---
title: "Pan-cancer quantification of neoantigen-mediated immunoediting in cancer evolution"
author: ["Tao Wu", "Xue-Song Liu (Corresponding author)"]
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    toc_depth: 3
    self_contained: false
---


```{r knitr_init, echo=FALSE}
library(knitr)
library(rmdformats)
## Global options
options(max.print = "75")
opts_chunk$set(
  echo = TRUE,
  comment = "#>",
  cache = TRUE, dpi = 300
)
opts_knit$set(width = 75)
Sys.setenv("LANGUAGE"="EN")
```

# Dependencies

This project is depended on R software, R packages, shell bash, Julia software and some Julia packages :

- [NeoEnrichment](https://github.com/wt12318/NeoEnrichment) - do neoantigen enrichment for this project.
- [tidyverse](https://www.tidyverse.org/) - tidy data
- [data.table](https://github.com/Rdatatable/data.table) - read and tidy data
- [survival](https://github.com/therneau/survival) - survival analysis 
- [survminer](https://github.com/kassambara/survminer) - plot survival fit
- [DT](https://cran.r-project.org/web/packages/DT/index.html) - show data table as a table in html
- [ggstatsplot](https://github.com/IndrajeetPatil/ggstatsplot) - plot scatter with linear fit
- [corrplot](https://cran.r-project.org/web/packages/corrplot/) - plot correlation
- [Julia](https://julialang.org/) - do simulation
- knitr, rmdformats - used to compile this file

# Data download and preprocessing

```{r child = 'data_download_preprocessing.Rmd'}
```

# TCGA pancancer analysis
```{r child = 'tcga_pancancer_analysis.Rmd'}
```

# Simulation analysis

# Supplementary analysis


